opXRD: Open Experimental Powder X-ray Diffraction Database

### 2025-03-06 chapters/01_introduction.tex
Updated state of the art ML

Both labeled and unlabeled datasets of experimental powder diffractograms hold significant value for machine learning-based pXRD analysis, particularly with regard to bridging the performance gap between simulated and experimental domains. Labeled experimental data can be used to test and benchmark existing and new automated analysis approaches. This enables researchers to gauge how well a given model would perform under real-world conditions if integrated into an automated experimentation pipeline. Unlabeled experimental data enables machine learning researchers to evaluate how closely their simulations represent experimental data and modify their simulation algorithms accordingly. Unlabeled data could also find applications in transfer learning approaches to transfer model capabilities from the domain of simulated diffractograms to the domain of experimental diffractograms. While some experimental powder databases exist, their utility is limited by the fact that they are either small or not openly accessible. \\

In this work, we introduce an open powder X-ray diffraction (opXRD) database featuring a broad range of patterns collected from experiments. With a total of $92,552$ patterns collected from 6 contributing institutions, the opXRD database exceeds the size of the previously largest database of openly accessible experimental powder diffraction data by two orders of magnitude. To the best of our knowledge, the largest database of this type is the RRUFF database, containing 1290 experimental powder diffraction patterns \cite{lafuente2015}. Further commercial datasets such as the PDF5+\cite{GatesRector2019} and the Linus Pauling File\cite{villars2018} exist, but license fees and restrictions in their use to train publishable models or publish the training data limit their usability. Fig.~$\eqref{fig:overview}$ provides an overview of machine learning workflows enabled and supported by the opXRD database. \\

% VISUAL ABSTRACT FIGURE
\begin{figure}[!htb]
    \centering
    \includegraphics[width=1.0\linewidth]{figures/pipeline.pdf}
    \caption{Experimental powder X-ray diffraction (pXRD) patterns from several contributors are collected in the \textit{opXRD} database. The proposed open-access database of experimental data aims to support each step in the pXRD-related machine learning workflow by informing better physics simulations, supplying model training data, and providing a foundation for realistic performance evaluations.}
    \label{fig:overview}
\end{figure}

@@

These imperfections include preferred crystallations orientation, variations in grain size, crystal defects, the impact of temperature on the scattering process, internal stress, the non-monochromacity of the X-ray source and X-ray-induced fluorescence\cite{cao2024simxrd, Waseda2011, Pecharsky2023}. Additionally, it should be considered that varying experimental setups can produce distinct powder diffraction patterns on the same sample. Features that may vary between experimental setups include the shape of diffraction peaks, the wavelength and polarization of the employed X-ray source, and the detector geometry \cite{cao2024simxrd, Waseda2011, Pecharsky2023}. The recorded scattering angles may also be slightly falsified if the sample is displaced from its intended position\cite{cao2024simxrd,hulbert2023}.

Additionally, some of the observed imperfections can be accounted for using transfer learning approaches. In transfer learning the objective is to transfer the capabilities of a model learned on a source domain in which labeled data is abundant to a target domain in which labeled data is sparse\cite{Zhuang2021}. In this context, the source domain is simulated powder diffraction patterns and the target domain is experimental powder diffraction patterns. Many approaches to transfer learning have been proposed, particularly in the domain of image classification \cite{Gatys2016, Ganin2015}. These existing techniques can be adapted to facilitate transfer learning in the context of pXRD patterns. Seddiki et al. have successfully applied transfer learning in the domain of mass spectrometry to boost the accuracy of mass spectrum classification models\cite{Seddiki2020}. Since both mass spectrometry data and pXRD data are one-dimensional, this work demonstrates the merit of transfer learning in a setting similar to pXRD. \\

Our opXRD dataset has been curated by collecting the accumulated data from multiple research groups and institutions with high-throughput XRD facilities. It contains pXRD patterns from single and multiphase materials from a wide variety of material classes, including high-entropy materials, perovskites, and commercial catalysts. Some of the XRD data was collected on thin-films rather than on true powder samples, which may influence the quality of the data in regards to full structure resolution. Additionally, some of the data was collected in grazing-angle geometry rather than in the usual Bragg-Brentano geometry employed in powder diffraction. Details about the experiments of contributing research groups, including the presence of thin-films samples and the usage of atypical diffraction geometries are discussed in Section~$\eqref{sec:our_dataset}$. A detailed description of how to use the data is given in Section~$\eqref{sec:how_to_use}$, and Section~$\eqref{sec:summary_and_outlook}$ describes how further data can be contributed. \\

\begin{figure*}[!htb]
    \centering
    \includegraphics[width=\linewidth]{figures/overview.pdf}
    \caption{Overview of the data collection pipeline. Datasets are submitted using an online submission form, optionally with the help of our submission helper software. After post-processing and data homogenization, we offer the creation of a Zenodo entry for each user submission and subsequently include the submission in the opXRD database.}
    \label{fig:overview}
\end{figure*}

% How to contribute? This aspect deserves a separate paragraph
The opXRD database is intended as a growing, community-driven initiative. The dataset we present here is the first version, but we hope to further increase the dataset size through active engagement with the pXRD community. Our primary objective is to minimize the effort and thus the barrier to contributing experimental data to the opXRD database. Thus, we developed a program that helps to find and share data from pXRD lab computers. In detail, users can select their most common pXRD file types, the program lists all files of that type, and users can select or deselect certain folders or files for sharing. Selected contributions will be uploaded to opXRD, processed to a common file format, and---if wanted---published on Zenodo on behalf of the contributors, before becoming part of the opXRD database. If labels are available, they can be shared with opXRD as well. Further details can be found on the opXRD website (\url{https://xrd.aimat.science/}).  \\

As argued by Aranda and Kroon-Batenburg \textit{et al}.\cite{Aranda2018, Kroon-Batenburg2024}, sharing raw powder diffraction data is not only in the interest of furthering machine learning research but is also in line with open science principles. It furthers the ability of other researchers to reproduce published work and in turn, adds to the credibility of the publisher of the data. Considering our streamlined submission and post-processing framework, a publication of powder diffraction data through the opXRD database is a simple way of realizing these principles. Compared to publishing data individually, it also has the added benefit that the data is contributed towards a large, homogenous dataset with a standardized interface. This makes the data more easily accessible to other researchers and provides more value to researchers seeking large quantities of data compared to an isolated submission on Zenodo. However, further data annotation with metadata is required to fully fulfill the FAIR data principles.\\

The broad range of available experimental samples contained in the opXRD v1.0 database makes it possible to apply state-of-the-art ML approaches to the domain of powder XRD analysis. We hope that this can drive ML research in this field towards more advanced automated analysis workflows that can accelerate materials science research through ready application in high-throughput experimentation pipelines.\\

>>>>

Both labeled and unlabeled datasets of experimental powder diffractograms hold significant value for machine learning-based pXRD analysis, particularly with regard to bridging the performance gap between simulated and experimental domains. Labeled experimental data can be used to test and benchmark existing and new automated analysis approaches. This enables researchers to gauge how well a given model would perform under real-world conditions if integrated into an automated experimentation pipeline. Unlabeled experimental data enables machine learning researchers to evaluate how closely their simulations represent experimental data and modify their simulation algorithms accordingly. Unlabeled data can also find applications in transfer learning approaches to transfer model capabilities from the domain of simulated diffractograms to the domain of experimental diffractograms. While some experimental powder databases exist, their utility is limited by the fact that they are either small or not openly accessible. \\

In this work, we introduce an open powder X-ray diffraction (opXRD) database featuring a broad range of patterns collected from experiments. With a total of $92,552$ patterns collected from 6 contributing institutions, the opXRD database exceeds the size of the previously largest database of openly accessible experimental powder diffraction data by two orders of magnitude. To the best of our knowledge, the largest database of this type is the RRUFF database, containing 1290 experimental powder diffraction patterns \cite{lafuente2015}. Larger commerical datasets such as the PDF5+\cite{GatesRector2019} and the Linus Pauling File\cite{villars2018} exist, but their utility is limited by fees and restrictive licenses. License terms of commercial datasets such as the PDF5+ and the Linus Pauling File prohibit or restrict the publication of models trained on their data. In contrast, the opXRD database is both free and imposes no restrictions on how its data is used. Fig.~$\eqref{fig:ml_uses}$ provides an overview of machine learning workflows enabled and supported by the opXRD database. \\

% VISUAL ABSTRACT FIGURE
\begin{figure}[!htb]
    \centering
    \includegraphics[width=1.0\linewidth]{figures/pipeline.pdf}
    \caption{Experimental powder X-ray diffraction (pXRD) patterns from several contributors are collected in the \textit{opXRD} database. The proposed open-access database of experimental data aims to support each step in the pXRD-related machine learning workflow by informing better physics simulations, supplying model training data, and providing a foundation for realistic performance evaluations.}
    \label{fig:ml_uses}
\end{figure}

@@

The neglected effects that lead to discrepancies between simulated patterns and patterns stemming from experiments are largely known. Unaccounted effects may include preferred crystallite orientation, variations in grain size, crystal defects, the impact of temperature on the scattering process, internal stress, the non-monochromaticity of the X-ray source, and X-ray-induced fluorescence\cite{cao2024simxrd, Waseda2011, Pecharsky2023}. Additionally, varying experimental setups produce distinct powder diffraction patterns on the same sample. Features that may vary between experimental setups include the shape of diffraction peaks, the wavelength and polarization of the employed X-ray source, and the detector geometry \cite{cao2024simxrd, Waseda2011, Pecharsky2023}. The recorded scattering angles may also be slightly falsified if the sample is displaced from its intended position\cite{cao2024simxrd,hulbert2023}. As these and more neglected effects are integrated into the simulation process, real powder diffraction data can be used to evaluate how closely simulated data matches up with real data. While direct comparisons are only possible on labeled patterns, comparing the strength and prevalence of features between simulated and real data can nevertheless provide clues about the fidelity of the situation. Taking on board all neglected effects without making approximations will incur significant computational costs that will lower the size of the generated training data. A more efficient approach could be to use real experimental data to identify the effects that have the largest impact in practice and model them heuristically. \\

The second way in which unlabeled experimental data can serve to bridge the performance gap between simulated and experimental domains is through transfer learning. The objective of transfer learning is to transfer the capabilities of a model learned on a source domain in which labeled data is abundant to a target domain in which labeled data is sparse\cite{Zhuang2021}. In this context, the source domain is simulated powder diffraction patterns and the target domain is experimental powder diffraction patterns. Many approaches to transfer learning have been proposed, particularly in the domain of image classification \cite{Gatys2016, Ganin2015}.  These existing techniques can be adapted to facilitate transfer learning in the context of pXRD patterns. Additionally, Seddiki et al. have already successfully applied transfer learning in the domain of mass spectrometry to boost the accuracy of mass spectrum classification models\cite{Seddiki2020}. Since both mass spectrometry data and pXRD data are one-dimensional, this work demonstrates the merit of transfer learning in a setting similar to pXRD. \\

% Contributing + community driven
The database is intended as a growing, community-driven initiative. The database we present here is the first version, but we hope to further increase the database size through active engagement with the pXRD community. Our primary objective is to minimize the effort and thus the barrier to contributing experimental data to the opXRD database. Thus, we developed a program that helps to find and share data from pXRD lab computers. Users can select their most common pXRD file types, the program lists all files of that type, and users can select or deselect certain folders or files for sharing. Selected contributions will be uploaded to opXRD, processed to a common file format, and---if wanted---published on Zenodo on behalf of the contributors, before becoming part of the opXRD database. If labels are available, they can be shared with opXRD as well. Further details can be found on the opXRD website (\url{https://xrd.aimat.science/}). An overview of this process is given in below Fig.~$\eqref{fig:submission}$.  \\

\begin{figure*}[!htb]
    \centering
    \includegraphics[width=\linewidth]{figures/overview.pdf}
    \caption{Overview of the data collection pipeline. Datasets are submitted using an online submission form, optionally with the help of our submission helper software. After post-processing and data homogenization, we offer the creation of a Zenodo entry for each user submission and subsequently include the submission in the opXRD database.}
    \label{fig:submission}
\end{figure*}

As argued by Aranda and Kroon-Batenburg \textit{et al}.\cite{Aranda2018, Kroon-Batenburg2024}, sharing raw powder diffraction data is not only in the interest of furthering machine learning research but is also in line with open science principles. It furthers the ability of other researchers to reproduce published work and in turn, adds to the credibility of the publisher of the data. Compared to publishing data individually, it also has the added benefit that the data is contributed towards a large, homogenous dataset with a standardized interface. This makes the data more easily accessible to other researchers and provides more value to researchers seeking large quantities of data compared to an isolated submission on Zenodo. However, further data annotation with metadata is required to fully fulfill the FAIR data principles.\\

The opXRD database contains pXRD patterns from single and multiphase materials from a wide variety of material classes, including high-entropy materials, perovskites, and commercial catalysts. Some of the XRD data was collected on thin-films rather than on true powder samples, which may influence the quality of the data in regards to full structure resolution. Additionally, some of the data was collected in grazing-angle geometry rather than in the usual Bragg-Brentano geometry employed in powder diffraction.
The broad range of available experimental samples contained in the opXRD v1.0 database makes it possible to apply state-of-the-art ML approaches to the domain of powder XRD analysis. We hope that the opXRD database can drive ML research in this field towards more advanced automated analysis workflows that can accelerate materials science research through ready application in high-throughput experimentation pipelines. Details of the experiments of research groups contributing to the opXRD database are discussed in Section~$\eqref{sec:our_dataset}$. A detailed description of how to acquire and use opXRD data is given in Section~$\eqref{sec:how_to_use}$, and Section~$\eqref{sec:summary_and_outlook}$ describes how further data can be contributed.

### 2025-03-06 chapters/02_SOTA_ML.tex
Updated state of the art ML

To showcase the need for datasets such as the one presented in this publication, we now shortly discuss some recent
approaches that apply machine learning methods to classification and regression tasks for powder diffractograms. \\

@@

The ultimate goal of ML efforts is to make pXRD data identification practical for experimental use, automating structure prediction. However, the current volume of experimental pXRD patterns is insufficient to effectively train ML models, highlighting an urgent need for a comprehensive experimental pXRD database. Two key approaches are essential to address this. First, developing more sophisticated simulation methods to better approximate experimental patterns\cite{cao2024simxrd} by using statistics from experimental diffractograms. Second, creating an experimental database that enables transfer learning to bridge the gap between simulated and real-world data. For both of these steps, the development of opXRD is particularly significant, as it will provide a comprehensive experimental benchmark for the community, allowing fair comparison of baseline models and accurate evaluation of their applicability in real experimental situations.

>>>>

\subsection*{Review of machine learning-based pXRD analysis}
To showcase the need for datasets such as the one presented in this publication, we now discuss some recent approaches that apply machine learning methods to classification and regression tasks for powder diffractograms. \\

@@

The current volume of experimental pXRD patterns is insufficient to effectively train ML models, highlighting an urgent need for a comprehensive experimental pXRD database. The most advanced ML models currently are trained on approximately $10^5 - 10^6$ simulated diffractograms. \cite{Salgado2023, Schopmans2023} This is, to the best of our knowledge, two orders of magnitude larger than the largest currently curated experimental dataset, the PDF-5+ with approximately $2\cdot 10^4$ experimental patterns. It is even one order of magnitude larger than the approximately $10^5$ unlabeled diffractograms in the initial version of the opXRD dataset we present here. \\

To make ML-based pXRD data identification practical for experimental use despite lacking experimental training data two key approaches are essential. First, developing more sophisticated simulation methods to better approximate experimental patterns\cite{cao2024simxrd} by using statistics from experimental diffractograms. Second, creating an experimental database that enables transfer learning to bridge the gap between simulated and real-world data. For both of these steps, the development of opXRD is particularly significant, as it will provide a comprehensive experimental benchmark for the community, allowing fair comparison of baseline models and accurate evaluation of their applicability in real experimental situations.

\pagebreak
